\MakeAbstract{
    Unix V6++ 是同济大学操作系统课程使用的教学内核，继承了 Unix V6 的进程、内存、文件和设备管理结构。原系统面向 IA32 架构并采用 C++ 实现，适合课程阅读和实验修改，但其内存管理、文件系统和缓冲区缓存路径仍大量依赖裸指针、全局状态和手工约定。随着教学平台向 Rust 和 RISC-V 迁移，原有实现在线程安全边界、资源生命周期表达和跨架构适配方面的维护成本逐渐增加。

    本文在团队整体 Rust 重构和 RISC-V 迁移工作的基础上，重点研究 Unix V6++ 的内存管理、文件系统及其相关 I/O 路径。内存管理部分重构了物理页表示、页级分配、内核堆和交换区管理，引入类型化地址抽象、Buddy 与 Slab 分层分配以及 RAII 风格的资源回收机制。文件系统部分保留超级块、inode、目录项、打开文件表和多级块映射等 Unix V6 风格结构，重新整理 inode、文件对象、文件描述符和缓冲区缓存之间的引用关系，并使用 \texttt{Result<T, E>} 和错误枚举组织错误传播路径。在 RISC-V 迁移方面，本文围绕内存管理和文件系统所依赖的底层条件，完成 Sv39 分页适配、MMIO UART 串口和 VirtIO 块设备路径分析，使文件系统上层逻辑能够在 IA32 与 RISC-V 两个平台上复用。

    本文通过原始 C++ IA32、Rust IA32 和 Rust RISC-V 三个版本的功能测试与性能测试验证重构结果。测试表明，两个 Rust 版本均能够完成文件系统装载、shell 运行、文件读写和用户程序执行等基本任务。系统调用、动态内存分配和文件 I/O 测试结果显示，Rust 重构没有引入不可接受的整体开销，部分文件系统相关系统调用在重构后延迟降低。总体而言，本文工作在保留 Unix V6++ 教学连续性的前提下，使内存管理和文件系统的资源生命周期、错误处理和平台边界表达更明确，为后续课程实验和体系结构迁移提供了基础。
}{Unix V6++，Rust，RISC-V，内存管理，文件系统，教学操作系统}

\MakeAbstractEng{
    Unix V6++ is a teaching operating system kernel used in the operating system course at Tongji University. It inherits the process, memory, file, and device management structure of Unix V6. The original system targets the IA32 architecture and is implemented in C++, which makes it suitable for source-code reading and laboratory exercises. However, the memory management, file system, and buffer cache paths still rely heavily on raw pointers, global states, and manually maintained conventions. As the teaching platform moves toward Rust and RISC-V, the original implementation shows increasing maintenance costs in safety boundaries, resource lifetime expression, and cross-architecture adaptation.

    Based on the team's overall Rust refactoring and RISC-V porting work, this thesis focuses on the memory management, file system, and related I/O paths of Unix V6++. For memory management, it refactors physical page representation, page-level allocation, kernel heap management, and swap-space management. Typed address abstractions, a layered Buddy and Slab allocation mechanism, and RAII-style resource reclamation are introduced. For the file system, it preserves the Unix V6 style structures such as the superblock, inodes, directory entries, open file tables, and multi-level block mapping, while reorganizing the reference relationships among inodes, file objects, file descriptors, and buffer cache entries. Error propagation is expressed through \texttt{Result<T, E>} and explicit error enumerations. For the RISC-V port, the thesis adapts the lower-level conditions required by memory management and the file system, including Sv39 paging, MMIO UART, and VirtIO block device paths, so that the upper file-system logic can be reused across IA32 and RISC-V platforms.

    The refactoring is evaluated through functional and performance tests on three versions: the original C++ IA32 version, the Rust IA32 version, and the Rust RISC-V version. The results show that both Rust versions can mount the file system, run the shell, perform file operations, and execute user programs. System call, dynamic memory allocation, and file I/O tests indicate that the Rust refactoring does not introduce unacceptable overall overhead, and some file-system-related system calls show lower latency after refactoring. In summary, this work preserves the teaching continuity of Unix V6++ while making resource lifetimes, error handling, and platform boundaries in memory management and the file system more explicit. It provides a basis for further course experiments and architecture migration.
}{Unix V6++, Rust, RISC-V, memory management, file system, teaching operating system}
